ODGI: scalable tools for pangenome graphs

### 2021-06-02 manuscript/main.tex
fixes

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are produced at a high rate, including
    efforts from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for homologs, paralogs and
    orthologs. So-called paths can represent reads, contigs, haplotypes,
    individual chromosomes or part thereof, that run through the
    nodes. A bi-directed graph can even contain both strands of DNA
    as paths. A graph-based approach is particularly relevant for
    exploring complex variation and highly repetitive regions present
    which characterises the vertebrate genomes.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but does not scale for
    large data (see Table~\ref{tab:02}). We
    wrote a new toolset and Application Programming Interface (API)
    in \Cpp\ named Optimized Dynamic Genome/Graph Implementation (ODGI).
    \odgi\ implements a efficient graph structure in computer memory
    that can be dynamically updated using multiple CPU cores in parallel.

@@

    % \vspace{-0.1in}

@@

@@

    \odgi\ is designed to build and modify paths in parallel. It
    applies concepts first introduced in the dynamic version of the
    graph extension of the positional Burrows-Wheeler
    transform~\citep{Siren:2020}. For implementation details see the
    \odgi\ online documentation. \odgi\ currently comes with over 30
    tools, including:

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    (GFAv1) format\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
    it validates paths with respect the graph topology.

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome (see Fig.~\ref{fig:1}a)
    and is capable of handling full length human chromosomes. \cmd{odgi sort}
    reorganises node order and simplifies the graph by applying several
    algorithms, including the novel 'path-guided stochastic gradient descent'
    algorithm. \cmd{odgi layout} generates a layout file with the X-Y
    coordinates for all nodes and \cmd{odgi draw} draws it, bringing
    the pangenome representation in two dimensions.

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into genome complexity in other
    ways. \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarises path
    information into bins of a specified size, thereby enabling a
    summarised view of large genomes. \cmd{odgi depth} and
    \cmd{odgi degree} return the node depth and degree as defined by
    user provided criteria. These methods allow the detection of complex
    genomic regions generated by highly repetitive sequences present in
    centromeres, segmental duplications, and acrocentric chromosomes.

>>>>

    Thanks to advances in sequencing technology, new full-length
    genome assemblies are produced at a high rate, including efforts
    from the Human Pangenome Reference Consortium (HPRC) and
    telomere-to-telemere projects, e.g.~\citep{Miga:2020}. In
    principle, a pangenome can model a full set of genomic elements of
    individuals in a given species or clade. Pangenomics contrasts
    with reference genome based approaches which relate sequences to a
    single consensus model of the genome~\citep{Eizenga:2020}. A
    pangenome can elegantly be represented by a graph data structure
    incorporating sequences as nodes and their connections as
    edges. These nodes are shared for identical sequences, such as
    homologs, paralogs and orthologs. So-called paths can represent
    reads, contigs, haplotypes, individual chromosomes or part
    thereof, that run through the nodes. A bi-directed graph can even
    contain both strands of DNA as paths. A graph-based approach is
    particularly relevant for exploring complex variation and highly
    repetitive regions present which characterises the vertebrate
    genomes.

    A major challenge is writing software that can deal with the sheer
    size of these graphs representing hundreds, and soon thousands, of
    human genomes. The \vg\ toolkit pioneered graph
    processing~\citep{vgtools,Eizenga:2020b}, but we found it does not
    necessarily scale for large data (see Table~\ref{tab:02}). We
    wrote a new toolset and Application Programming Interface (API) in
    \Cpp\ named Optimized Dynamic Genome/Graph Implementation (ODGI).
    \odgi\ implements a efficient graph structure in computer memory
    that can be dynamically updated using multiple CPU cores in
    parallel.

@@

@@

    \vspace{-0.1in}

@@

    \odgi\ is designed to build and modify paths in parallel. It
    applies concepts first introduced in the dynamic version of the
    graph extension of the positional Burrows-Wheeler
    transform~\citep{Siren:2020}. For implementation details see the
    \odgi\ online documentation. \odgi\ currently comes with more than 30
    tools and here we present a subset of functionality:

    \indent \topic{odgi build, view \& validate} pangenome graphs can be
    stored in the textual Graphical Fragment Assembly
    format~\citep{GFA}. \cmd{odgi build} and \cmd{odgi view}
    convert graphs from GFA to the \odgi\ format and \textit{vice versa}.
    \cmd{odgi validate} ascertains the graph data is correct, e.g.,
    it validates paths with respect to the graph topology.

    \topic{odgi viz, sort, draw \& layout} pangenome visualisation
    provides convenient insight into genomic variation. \cmd{odgi viz}
    generates a linearized representation of the pangenome (see Fig.~\ref{fig:1}a)
    and is capable of handling full length human chromosomes. \cmd{odgi sort}
    reorganises node order and simplifies the graph by applying several
    algorithms, including our novel 'path-guided stochastic gradient descent'
    algorithm. \cmd{odgi layout} generates a layout file with the X-Y
    coordinates for all nodes and \cmd{odgi draw} draws it, bringing
    the pangenome representation in two dimensions.

    \topic{odgi stats, bin, depth \& degree} pangenome numerical
    counts provide insight into complexity in other
    ways. \cmd{odgi stats} returns the number of nodes,
    edges, paths, and graph length. \cmd{odgi bin} summarises path
    information into bins of a specified size, thereby enabling a
    summarised view of large genomes. \cmd{odgi depth} and
    \cmd{odgi degree} return the node depth and degree as defined by
    user provided criteria. These methods allow the detection of complex
    genomic regions generated by highly repetitive sequences.
